Impact of IPTp Regimen on Pregnancy Outcomes in a Malaria-Endemic Setting

### 2025-04-28 code/analysis-code/Statistical_Analysis.qmd
Update

# Data manipulation and wrangling
library(dplyr)              # Data manipulation
library(tidyverse)          # Core tidyverse packages (ggplot2, readr, etc.)
library(janitor)            # Clean column names
library(skimr)              # Quick data summaries
library(lubridate)          # Date/time handling
library(forcats)            # Categorical variable tools

# Visualization
library(ggplot2)            # Data visualization
library(ggpubr)             # Publication-ready plots

@@

@@

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, parsnip, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine
library(generalhoslem)
library(ResourceSelection)
library(rms)
library(car)

@@

```{r}
# Data Import & Initial Inspection 
# load the Dataset
PROMO_Data <- read_csv(here("data", "raw-data", "PROMO_Data.csv"))

promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))
```

@@

```{r}
names(promo_data_clean)
```

*Table 1*

```{r}
# Recode variables for baseline table presentation:
promo_data_clean <- promo_data_clean %>%
  mutate(
    # Create an age group variable (for potential subgroup analyses)
    age_group = ifelse(age_at_enrollment_years < 25, "Young", "Older"),

@@

```{r compare-models-lrt-vif, message=FALSE, warning=FALSE}
# 1) Subset to the exact variables and drop any rows with missingness
analysis_data <- promo_data_clean %>%
  select(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  ) %>%
  drop_na()

@@

# 5) Tidy LRT into a data frame
anova_df <- as.data.frame(anova_out) %>%
  mutate(
    Model = c("Model 1 (No Interaction)", 
              "Model 2 (With Interaction)")
  ) %>%
  select(
    Model,
    `Resid. Df`,
    `Resid. Dev`,
    Df,
    Deviance,
    `Pr(>Chi)`
  )

@@

# 2) Compute VIFs treating interaction terms as separate predictors
vif_mat <- vif(model2, type = "predictor")

# 3) Tidy into a data frame
vif_df <- as.data.frame(vif_mat) %>%
  rownames_to_column("Variable") %>%
  rename(
    GVIF           = GVIF,
    `Df (VIF)`     = Df,
    `GVIF^(1/(2*Df))` = `GVIF^(1/(2*Df))`
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "total_malaria_episodes"              ~ "Total malaria episodes",
      Variable == "study_armSP"                         ~ "Treatment arm (SP vs DP)",
      Variable == "age_at_enrollment_years"             ~ "Age at enrollment",
      Variable == "gravidity"                           ~ "Gravidity",
      grepl("^education_level", Variable)               ~ 
        paste("Education:", sub("education_level", "", Variable)),
      Variable == "total_malaria_episodes:study_armSP"  ~ "Interaction: episodes × SP",
      TRUE                                              ~ Variable
    )
  ) %>%
  select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`)

# 4) Print the VIF table to the console
print(vif_df)

@@

## Random Forest Model Specification and Hyperparameter Tuning
*Train/Test Split and Random Forest Model* which fits a random forest model and plots its ROC curve, is also used for Research Question 1 as an alternative ML approach
```{r}
# 0) Make sure the outcome is a factor
train_data <- train_data %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0, 1)))

# 1) Define the random‐forest spec
rf_model <- rand_forest(
  trees = 500,
  mtry  = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

library(dials)   # for finalize()

# 2) Build the recipe 
rf_recipe <- recipe(adverse_birth_outcome ~ 
                      total_malaria_episodes + study_arm + 
                      age_at_enrollment_years + gravidity + education_level,
                    data = train_data) %>%
  step_dummy(all_nominal_predictors())

# 3) Assemble workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

# 4) Extract and finalize parameters
rf_params <- parameters(rf_model) %>%
  finalize(train_data)    # now mtry knows how many predictors

# 5) Create a random tuning grid of size 20
rf_grid <- grid_random(rf_params, size = 20)

# 6) 10-fold stratified CV
rf_folds <- vfold_cv(train_data, v = 10, strata = adverse_birth_outcome)

# 7) Tune
set.seed(123)
rf_tune <- tune_grid(
  rf_workflow,
  resamples = rf_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc)
)

# 8) View results
rf_tune %>% 
  collect_metrics() %>% 
  filter(.metric == "roc_auc")

# Random Forest model training and OOB Brier score output
```{r}
# 1)  outcome is a factor
train_data <- train_data %>%
  mutate(
    adverse_birth_outcome = factor(
      adverse_birth_outcome,
      levels = c(0, 1),
      labels = c("no", "yes")
    )
  )

# 2) Define a Random Forest model (with fixed hyperparameters for illustration)
rf_model <- rand_forest(
    trees = 500,
    mtry  = 3,    # you can choose a sensible default or tune later
    min_n = 5     # likewise
  ) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# 3) Build the recipe
rf_recipe <- recipe(
    adverse_birth_outcome ~
      total_malaria_episodes +
      study_arm +
      age_at_enrollment_years +
      gravidity +
      education_level,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors())

# 4) Assemble into a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rf_recipe)

# 5) Fit the random forest
rf_fit <- rf_workflow %>% 
  fit(data = train_data)

# 6) Inspect the fitted workflow
rf_fit

```
- Random forest fit ran without errors, but the OOB Brier score shows it’s essentially no better than random. 
- That echoes what you saw with your logistic and other ML models (AUCs ~0.4–0.5): with these variables alone, none of the algorithms can predict the composite adverse birth outcome.

@@

I started by performing a 70/30 train/test split, stratified on the adverse birth outcome. For Research Question 1, I fit a logistic regression model with an interaction term on the training set, generated predictions on the test set, and plotted the ROC curve (with an AUC computed to assess discriminative performance). Then, I fit a random forest classifier (converting factor predictors to dummies in the recipe) and similarly evaluated its performance on the test set using ROC analysis. Finally, I overlaid both ROC curves to compare the models directly. This approach provides an "honest" assessment of model performance on unseen data and allows me to explore additional machine learning methods for addressing my research question.

@@

# Define a boosting model for classification (using boost_tree)
```{r}
# Define a boosting model for classification (using boost_tree)
boost_model <- boost_tree(
  mode = "classification",
  trees = 500,
  learn_rate = tune(),
  tree_depth = tune()
) %>%
  set_engine("xgboost")

# Build a recipe (convert factor predictors to dummy variables)
boost_recipe <- recipe(adverse_birth_outcome ~ total_malaria_episodes + study_arm +
                         age_at_enrollment_years + gravidity + education_level,
                       data = train_data) %>%
  step_dummy(all_nominal_predictors())

# Create a workflow
boost_workflow <- workflow() %>%
  add_model(boost_model) %>%
  add_recipe(boost_recipe)

# Create a tuning grid for learning rate and tree depth
boost_grid <- grid_regular(
  learn_rate(range = c(0.01, 0.3)),
  tree_depth(range = c(3, 8)),
  levels = 5
)

# Create 10-fold cross-validation folds from the training data
cv_folds <- vfold_cv(train_data, v = 10)

# Perform 10-fold cross-validation tuning using the boosting model workflow
boost_tune <- tune_grid(
  boost_workflow,
  resamples = cv_folds,
  grid = boost_grid
)

# Select best parameters based on ROC AUC
best_boost <- select_best(boost_tune, metric = "roc_auc")

# Finalize the workflow with best parameters
final_boost <- finalize_workflow(boost_workflow, best_boost)

# Fit the final boosting model on the training data
boost_fit <- final_boost %>% fit(data = train_data)

# Generate predicted probabilities on the test set
boost_preds <- predict(boost_fit, new_data = test_data, type = "prob")

# Compute ROC for the boosting model using the probability for class "1"
roc_boost <- roc(test_data$adverse_birth_outcome, boost_preds$.pred_1)
plot(roc_boost, col = "green", lwd = 2, main = "ROC Curve: Boosting Model")
auc_boost <- auc(roc_boost)
print(paste("Boosting Model AUC:", round(auc_boost, 2)))

The ROC curve for the boosting model rises more clearly above the diagonal line than the previous models, suggesting it does a better job distinguishing between those who experience adverse birth outcomes and those who do not. Although it’s still not perfect, the curve indicates an improvement in predictive performance compared to the logistic regression and random forest models, meaning the boosting model likely has a higher AUC and provides a more accurate risk estimate for adverse outcomes in this dataset.

@@

```{r}
# Fit logistic regression with rms
dd <- datadist(train_data); options(datadist = "dd")
rms_model <- lrm(adverse_birth_outcome ~ gravidity + total_malaria_episodes + study_arm + education_level,
                 data = train_data, x = TRUE, y = TRUE)
# Calibration plot
cal <- calibrate(rms_model, method="boot", B = 200)
plot(cal, xlab="Predicted probability", ylab="Observed proportion")

@@

```{r}
library(dplyr); library(ggplot2)

# 1) Get predictions on the train set
train_data <- train_data %>%
  mutate(
    pred = predict(model_gravidity, type = "response"),
    decile = ntile(pred, 10)
  )

# 2) Summarize observed vs. predicted in each decile
calib_df <- train_data %>%
  group_by(decile) %>%
  summarize(
    mean_pred = mean(pred),
    obs_rate  = mean(as.numeric(adverse_birth_outcome) - 1)
  )

# 3) Plot
ggplot(calib_df, aes(x = mean_pred, y = obs_rate)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    x = "Mean predicted probability",
    y = "Observed event rate",
    title = "Calibration plot (deciles)"
  ) +
  theme_minimal()

@@

```{r}
library(dplyr)
library(knitr)

enet_tune %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  slice_head(n = 5) %>%                # ← pull the top 5 rows
  mutate(`AUC (mean ± SE)` = sprintf("%.3f \u00B1 %.3f", mean, std_err)) %>%
  select(penalty, mixture, `AUC (mean ± SE)`) %>%
  kable(
    caption = "Table X. Top 5 elastic‐net hyperparameter combinations, ranked by mean cross‐validated AUC"
  )

@@

```{r}
enet_tune %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  head(5) %>%                          # ← base‐R head()
  mutate(`AUC (mean ± SE)` = sprintf("%.3f \u00B1 %.3f", mean, std_err)) %>%
  select(penalty, mixture, `AUC (mean ± SE)`) %>%
  kable(
    caption = "Table X. Top 5 elastic‐net hyperparameter combinations, ranked by mean cross‐validated AUC"
  )

@@

```{r}
# 2) Specify elastic‑net logistic regression
enet_spec <- logistic_reg(
    penalty = tune(),   # strength of regularization
    mixture = tune()    # 0=ridge ↔ 1=lasso
  ) %>%
  set_engine("glmnet")

@@

```{r}
# 3) Build the workflow
enet_wf <- workflow() %>%
  add_recipe(improved_recipe) %>%
  add_model(enet_spec)

```{r}
# 4) Create a tuning grid & 5) tune via 10‑fold CV
enet_grid <- grid_regular(
  penalty(range = c(-5, -1), trans = log10_trans()),  # 1e-5 → 1e-1
  mixture(range = c(0, 1)),
  levels = 5
)

```{r}
# 6) Select best, finalize, fit on train, & evaluate on test
best_enet  <- select_best(enet_tune, metric = "roc_auc")
enet_final <- finalize_workflow(enet_wf, best_enet)
enet_fit   <- fit(enet_final, data = train_data)

enet_preds <- predict(enet_fit, test_data, type = "prob") %>% bind_cols(test_data)
enet_auc   <- roc_auc(enet_preds, truth = adverse_birth_outcome, .pred_1)
print(enet_auc)

```
The elastic‑net model achieved an AUC of 0.49 on the held‑out test set—below the 0.50 threshold for random guessing. This indicates that, even with polynomial, interaction, and SMOTE steps, the model cannot reliably distinguish between pregnancies with and without adverse outcomes. Such near‑random performance suggests either insufficient predictive signal in the available features or the need for additional, more informative variables and further model refinement.

```{r}
# Create temporal & composite features in the training set
train_data <- train_data %>%
  mutate(
    enroll_month = lubridate::month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3))
  )

test_data <- test_data %>%
  mutate(
    enroll_month = lubridate::month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes / (gestational_age_at_enrollment_weeks / (40/3))
  )

```{r}
imbalanced_recipe <- recipe(adverse_birth_outcome ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_mutate(
    total_malaria_episodes = as.numeric(total_malaria_episodes),
    gravidity = as.numeric(gravidity)
  ) %>%
  step_adasyn(adverse_birth_outcome) %>%
  step_normalize(all_predictors())

```{r}
enet_wf <- workflow() %>%
  add_model(enet_spec) %>%
  add_recipe(imbalanced_recipe)

enet_bayes <- tune_bayes(
  enet_wf,
  resamples = cv_folds,
  initial = enet_tune,
  iter = 20,
  metrics = metric_set(roc_auc)
)

best_enet_bayes <- select_best(enet_bayes, metric = "roc_auc")

```{r}
library(stacks)

# Assume you've tuned logistic, rf, xgb already: logistic_tune, rf_tune, xgb_tune
stack <- stacks() %>%
  add_candidates(logistic_tune) %>%
  add_candidates(rf_tune)       %>%
  add_candidates(xgb_tune)      %>%
  blend_predictions()           %>%
  fit_members()

# Evaluate on test set
stack_preds <- predict(stack, test_data, type = "prob") %>% bind_cols(test_data)
stack_auc   <- roc_auc(stack_preds, truth = adverse_birth_outcome, .pred_1)
print(stack_auc)

```{r}
svm_spec <- svm_rbf(
    cost      = tune(),
    rbf_sigma = tune()
  ) %>%
  set_engine("kernlab") %>%
  set_mode("classification")

svm_wf <- workflow() %>%
  add_model(svm_spec) %>%
  add_recipe(improved_recipe)  # or your imbalanced_recipe

svm_grid <- grid_regular(
  cost(range = c(-5, 2), trans = log2_trans()),
  rbf_sigma(range = c(-3, 0), trans = log2_trans()),
  levels = 5
)

svm_tune <- tune_grid(
  svm_wf,
  resamples = cv_folds,
  grid      = svm_grid,
  metrics   = metric_set(roc_auc)
)

best_svm <- select_best(svm_tune, metric = "roc_auc")

```{r}
# For your best model (e.g. enet_fit or stack)
pr    <- pr_curve(stack_preds, truth = adverse_birth_outcome, .pred_1)
pr_auc <- pr_auc(stack_preds, truth = adverse_birth_outcome, .pred_1)

autoplot(pr) +
  labs(title = "Precision–Recall Curve", subtitle = paste("PR AUC =", round(pr_auc$.estimate,2)))

>>>>

# Tidyverse core for data import, manipulation & visualization
library(tidyverse)          # includes ggplot2, dplyr, tidyr, readr, purrr, etc.

# Data cleaning & summaries
library(janitor)            # Clean column names
library(skimr)              # Quick data summaries

# Date/time and categorical helpers
library(lubridate)          # Date/time handling
library(forcats)            # Categorical variable tools

@@

# Visualization enhancements
library(ggpubr)             # Publication-ready plots

@@

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine

# Goodness-of-fit & diagnostics
library(generalhoslem)      # Hosmer–Lemeshow tests
library(ResourceSelection)  # Additional GOF tests
library(rms)                # Regression modeling strategies
library(car)                # Companion to applied regression

@@

```{r}
# Data Import & Initial Inspection 
# load the Dataset
# ────────────────────────────────────────────────────────────────────────────────
# Project setup: anchor here() at the project root
# ────────────────────────────────────────────────────────────────────────────────
library(here)
here::i_am("data/raw-data/PROMO_Data.csv")

# ────────────────────────────────────────────────────────────────────────────────
# Load data using project‐relative paths
# ────────────────────────────────────────────────────────────────────────────────
library(readr)

PROMO_Data <- read_csv(
  here("data", "raw-data", "PROMO_Data.csv")
)

promo_data_clean <- read_csv(
  here("data", "clean", "PROMO_Data_clean.csv")
)
```

@@

*Table 1*
```{r}
# Recode variables for baseline table presentation:
promo_data_clean <- promo_data_clean %>%
  mutate(
    # Create an age group variable (for potential subgroup analyses)
    age_group = ifelse(age_at_enrollment_years < 25, "Young", "Older"),

@@

```{r compare-models-lrt-vif, message=FALSE, warning=FALSE}
# 1) Subset to the exact variables and drop any rows with missingness
analysis_data <- promo_data_clean %>%
  dplyr::select(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  ) %>%
  tidyr::drop_na()

@@

# 5) Tidy LRT into a data frame
anova_df <- as.data.frame(anova_out) %>%
  dplyr::mutate(
    Model = c("Model 1 (No Interaction)", "Model 2 (With Interaction)")
  ) %>%
  dplyr::select(
    Model,
    `Resid. Df`,
    `Resid. Dev`,
    Df,
    Deviance,
    `Pr(>Chi)`
  )

@@

# 2) Compute VIFs treating interaction terms as separate predictors
vif_mat <- car::vif(model2, type = "predictor")

# 3) Tidy into a data frame, explicitly using tibble::rownames_to_column() and dplyr:: verbs
vif_df <- as.data.frame(vif_mat) %>%
  tibble::rownames_to_column("Variable") %>%
  dplyr::rename(
    GVIF               = GVIF,
    `Df (VIF)`         = Df,
    `GVIF^(1/(2*Df))`  = `GVIF^(1/(2*Df))`
  ) %>%
  dplyr::mutate(
    Variable = dplyr::case_when(
      Variable == "total_malaria_episodes"             ~ "Total malaria episodes",
      Variable == "study_armSP"                        ~ "Treatment arm (SP vs DP)",
      Variable == "age_at_enrollment_years"            ~ "Age at enrollment",
      Variable == "gravidity"                          ~ "Gravidity",
      grepl("^education_level", Variable)              ~ paste0("Education:", sub("education_level", "", Variable)),
      Variable == "total_malaria_episodes:study_armSP" ~ "Interaction: episodes × SP",
      TRUE                                             ~ Variable
    )
  ) %>%
  dplyr::select(Variable, GVIF, `Df (VIF)`, `GVIF^(1/(2*Df))`)

# 4) View
print(vif_df)

@@

## Random Forest: Specification, Tuning & ROC Comparison
This section performs a 70/30 stratified train/test split, tunes a 500-tree random forest via 10-fold cross-validation to maximize ROC-AUC, fits the final model on the training set, and plots its ROC curve (alongside the logistic regression) as an alternative ML approach for Research Question 1.
```{r model-comparison, echo=TRUE}
# 0) Load required packages
library(dplyr)
library(rsample)     # for initial_split()
library(tidymodels)  # recipes, workflows, tuning
library(dials)       # for parameter grids
library(pROC)        # for ROC curves

# 1) Train/test split (70/30 stratified on outcome)
set.seed(1234)
split_data <- initial_split(promo_data_clean, prop = 0.7, strata = adverse_birth_outcome)
train_data <- training(split_data)
test_data  <- testing(split_data)

# 2) Convert outcome to a two‐level factor (no/yes)
train_data <- train_data %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1), labels = c("no","yes")))
test_data <- test_data %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1), labels = c("no","yes")))

# --------------------------------------------------
# 3) Fit logistic regression and build its ROC curve
# --------------------------------------------------
logit_fit <- glm(
  adverse_birth_outcome ~ 
    total_malaria_episodes * study_arm + 
    age_at_enrollment_years + 
    gravidity + 
    education_level,
  family = binomial(link = "logit"),
  data   = train_data
)

# Predict and construct ROC for logistic model
logit_probs <- predict(logit_fit, newdata = test_data, type = "response")
roc_logit  <- roc(test_data$adverse_birth_outcome, logit_probs, levels = c("no","yes"), direction = "<")

# --------------------------------------------------
# 4) Hyperparameter tuning for Random Forest
# --------------------------------------------------
rf_spec <- rand_forest(
    trees = 500,
    mtry  = tune(),
    min_n = tune()
  ) %>%
  set_engine("ranger") %>%
  set_mode("classification")

rf_recipe <- recipe(
    adverse_birth_outcome ~ 
      total_malaria_episodes + 
      study_arm + 
      age_at_enrollment_years + 
      gravidity + 
      education_level,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors())

rf_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(rf_recipe)

rf_params <- parameters(rf_spec) %>%
  finalize(train_data)

set.seed(123)
rf_tune <- tune_grid(
  rf_wf,
  resamples = vfold_cv(train_data, v = 10, strata = adverse_birth_outcome),
  grid      = grid_random(rf_params, size = 20),
  metrics   = metric_set(roc_auc)
)

# Select best hyperparameters and finalize
best_rf <- select_best(rf_tune, metric = "roc_auc")
rf_final_spec <- finalize_model(rf_spec, best_rf)

# --------------------------------------------------
# 5) Fit final Random Forest and build its ROC curve
# --------------------------------------------------
rf_final_wf <- workflow() %>%
  add_model(rf_final_spec) %>%
  add_recipe(rf_recipe)

rf_fit <- fit(rf_final_wf, data = train_data)

rf_probs <- predict(rf_fit, new_data = test_data, type = "prob") %>%
  pull(.pred_yes)

roc_rf <- roc(test_data$adverse_birth_outcome, rf_probs, levels = c("no","yes"), direction = "<")

# --------------------------------------------------
# 6) Plot both ROC curves together
# --------------------------------------------------
plot(
  roc_logit,
  col  = "blue",
  lwd  = 2,
  main = "ROC Comparison: Logistic Regression vs. Random Forest"
)
lines(roc_rf, col = "red", lwd = 2)
legend(
  "bottomright",
  legend = c("Logistic Regression", "Random Forest"),
  col    = c("blue", "red"),
  lwd    = 2
)
```

@@

I performed a 70/30 train/test split stratified on adverse birth outcome. On the training set I fit a logistic regression (including our pre-specified interaction term), then generated test-set predictions and computed its ROC curve (AUC = 0.57). In parallel, I tuned a 500-tree random forest via 10-fold CV (stratified on outcome), fit the final model, and similarly evaluated its ROC (AUC = 0.54). Finally, I overlaid both ROC curves against the 45° no-skill line. Both models lie only slightly above that baseline, indicating poor discrimination with our current predictors—though the logistic regression edges out the random forest by a slim margin.

@@

# Define a boosting model for classification (using boost_tree)

```{r boost-model, echo=TRUE}
library(dplyr)
library(rsample)
library(tidymodels)  # recipes, workflows, tune_grid
library(pROC)

# 0) Subset & drop any rows with missing outcome/predictors
ml_data <- promo_data_clean %>%
  dplyr::select(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  ) %>%
  tidyr::drop_na()

# 1) 70/30 split, stratified on the now‐clean outcome
set.seed(1234)
split_obj  <- initial_split(ml_data, prop = 0.7, strata = adverse_birth_outcome)
train_data  <- training(split_obj)
test_data   <- testing(split_obj)

# 2) Make sure the outcome is a 2‐level factor
train_data <- train_data %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1), labels = c("no","yes")))
test_data  <- test_data  %>%
  mutate(adverse_birth_outcome = factor(adverse_birth_outcome, levels = c(0,1), labels = c("no","yes")))

# 3) Boosting spec + recipe + workflow
boost_spec <- boost_tree(
    mode       = "classification",
    trees      = 500,
    learn_rate = tune(),
    tree_depth = tune()
  ) %>%
  set_engine("xgboost")

boost_rec <- recipe(
    adverse_birth_outcome ~
      total_malaria_episodes +
      study_arm +
      age_at_enrollment_years +
      gravidity +
      education_level,
    data = train_data
  ) %>%
  step_dummy(all_nominal_predictors())

boost_wf <- workflow() %>%
  add_model(boost_spec) %>%
  add_recipe(boost_rec)

# 4) Tuning grid + CV + tune_grid (with roc_auc)
boost_grid <- grid_regular(
  learn_rate(range = c(0.01, 0.3)),
  tree_depth(range = c(3, 8)),
  levels = 5
)

cv_folds <- vfold_cv(train_data, v = 10, strata = adverse_birth_outcome)

set.seed(2345)
boost_tune <- tune_grid(
  boost_wf,
  resamples = cv_folds,
  grid      = boost_grid,
  metrics   = metric_set(roc_auc)
)

# 5) Finalize, fit & evaluate
best_boost <- select_best(boost_tune, metric = "roc_auc")
final_boost_spec <- finalize_model(boost_spec, best_boost)

boost_final_wf <- workflow() %>%
  add_model(final_boost_spec) %>%
  add_recipe(boost_rec)

boost_fit <- fit(boost_final_wf, data = train_data)

# ROC on the test set
boost_probs <- predict(boost_fit, new_data = test_data, type = "prob") %>% pull(.pred_yes)
roc_boost  <- roc(test_data$adverse_birth_outcome, boost_probs, levels = c("no","yes"))
plot(roc_boost, col = "darkgreen", lwd = 2, main = "ROC Curve: XGBoost")
auc_boost <- auc(roc_boost)
print(paste("Boosting Model AUC:", round(auc_boost, 2)))
```
The ROC curve for the boosting model rises more clearly above the diagonal line than the previous models, suggesting it does a better job distinguishing between those who experience adverse birth outcomes and those who do not. Although it’s still not perfect, the curve indicates an improvement in predictive performance compared to the logistic regression and random forest models, meaning the boosting model likely has a higher AUC and provides a more accurate risk estimate for adverse outcomes in this dataset.

@@

```{r}
# only these variables are needed for the logistic model
dd <- datadist(
  train_data[, c(
    "gravidity",
    "total_malaria_episodes",
    "study_arm",
    "education_level",
    "adverse_birth_outcome"
  )]
)
options(datadist = "dd")

# now fit with rms
rms_model <- lrm(
  adverse_birth_outcome ~ 
    gravidity + total_malaria_episodes + study_arm + education_level,
  data = train_data,
  x = TRUE, y = TRUE
)

# bootstrap‐calibrated calibration curve
cal <- calibrate(rms_model, method = "boot", B = 200)
plot(
  cal,
  xlab = "Predicted probability",
  ylab = "Observed proportion"
)

@@

```{r}
library(dplyr)
library(ggplot2)

# 1) Get predictions on the train set
train_data <- train_data %>%
  mutate(
    pred   = predict(model_gravidity, newdata = ., type = "response"),
    decile = ntile(pred, 10)
  )

# 2) Summarize observed vs. predicted in each decile (explicitly using dplyr)
calib_df <- train_data %>%
  dplyr::group_by(decile) %>%
  dplyr::summarize(
    mean_pred = mean(pred, na.rm = TRUE),
    obs_rate  = mean(as.numeric(adverse_birth_outcome) - 1, na.rm = TRUE)
  )

# 3) Plot calibration curve
ggplot(calib_df, aes(x = mean_pred, y = obs_rate)) +
  geom_point(size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    x     = "Mean predicted probability",
    y     = "Observed event rate",
    title = "Calibration plot (deciles)"
  ) +
  theme_minimal()

@@

```{r}
library(dplyr)

# Get top 5 elastic‐net hyperparameter combinations and print as a regular tibble
top5_enet <- enet_tune %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  slice_head(n = 5) %>%
  mutate(`AUC (mean ± SE)` = sprintf("%.3f ± %.3f", mean, std_err)) %>%
  dplyr::select(penalty, mixture, `AUC (mean ± SE)`)

# Print the table
print(top5_enet)

@@

```{r}
library(dplyr)

# Extract and print the top 5 elastic-net tuning results as a regular tibble
top5_enet <- enet_tune %>%
  collect_metrics() %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  slice_head(n = 5) %>%
  mutate(`AUC (mean ± SE)` = sprintf("%.3f ± %.3f", mean, std_err)) %>%
  dplyr::select(penalty, mixture, `AUC (mean ± SE)`)

print(top5_enet)

@@

## Elastic Net Logistic Regression: Specification, Tuning & Evaluation

```{r feature-engineering}

library(dplyr)
library(tidyr)
library(lubridate)
library(rsample)
library(tidymodels)

# 0) Engineer features and impute their NAs
full_data <- promo_data_clean %>%
  mutate(
    enroll_month = month(enrollment_date),
    episodes_per_trimester = total_malaria_episodes /
      (gestational_age_at_enrollment_weeks / (40/3))
  )

# compute medians for imputation
median_enroll <- median(full_data$enroll_month, na.rm = TRUE)
median_eps    <- median(full_data$episodes_per_trimester, na.rm = TRUE)

full_data <- full_data %>%
  mutate(
    enroll_month           = if_else(is.na(enroll_month),    median_enroll,    enroll_month),
    episodes_per_trimester = if_else(is.na(episodes_per_trimester),
                                     median_eps,
                                     episodes_per_trimester)
  )

# 1) Safe stratification check
stratum_counts <- full_data %>% count(adverse_birth_outcome)
can_stratify   <- nrow(stratum_counts) >= 2 && all(stratum_counts$n >= 2)

# 2) 70/30 split (stratified if safe)
set.seed(1234)
split_obj <- if (can_stratify) {
  initial_split(full_data, prop = 0.7, strata = adverse_birth_outcome)
} else {
  warning("Not enough data to stratify; performing unstratified split.")
  initial_split(full_data, prop = 0.7)
}

raw_train <- training(split_obj)
raw_test  <- testing(split_obj)

# 3) Drop rows with NAs in the *core* six predictors or outcome
train_data <- raw_train %>%
  tidyr::drop_na(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  )

test_data <- raw_test %>%
  tidyr::drop_na(
    adverse_birth_outcome,
    total_malaria_episodes,
    study_arm,
    age_at_enrollment_years,
    gravidity,
    education_level
  )

# 4) Build dynamic v-fold CV (up to 10, but ≤ nrow(train_data))
n_train <- nrow(train_data)
if (n_train < 2) stop("Not enough rows for cross‐validation (need ≥ 2).")
v <- min(10, n_train)

cv_folds <- if (can_stratify) {
  vfold_cv(train_data, v = v, strata = adverse_birth_outcome)
} else {
  vfold_cv(train_data, v = v)
}

# 5) Finally, tune your workflow—no more stray NAs in enroll_month
enet_tune <- tune_grid(
  enet_wf,
  resamples = cv_folds,
  grid      = enet_grid,
  metrics   = metric_set(roc_auc)
)

@@

```{r}
library(tidymodels)

# (A) Tune hyperparameters via grid search
enet_tune <- tune_grid(
  enet_wf,               # your workflow
  resamples = cv_folds,  # the vfold_cv object you just built
  grid      = enet_grid, # the grid of penalty/mixture values
  metrics   = metric_set(roc_auc)
)

# (B) Look at the tuning results
collect_metrics(enet_tune) %>% 
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean))

# (C) Pick the best parameters
best_params <- select_best(enet_tune, "roc_auc")

# (D) Finalize & refit on all of train_data
final_enet_wf <- finalize_workflow(enet_wf, best_params)
enet_fit      <- fit(final_enet_wf, data = train_data)

# (E) Evaluate on the test set
enet_preds <- predict(enet_fit, new_data = test_data, type = "prob") %>%
  bind_cols(test_data)

enet_auc <- roc_auc(enet_preds, truth = adverse_birth_outcome, .pred_yes)
print(enet_auc)

### 2025-04-28 products/manuscript/Manuscript.qmd
P3
Update

# 2) Set your global chunk options (no setwd or root.dir needed):
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load Required Libraries for Data Handling, Visualization, and Analysis

# Load Required Libraries

@@

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, parsnip, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine
library(generalhoslem)
library(ResourceSelection)
library(rms)
```

>>>>

# 2) Set your global chunk options (no setwd or root.dir needed):
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold"}
# Load Required Libraries

@@

# Machine learning & modeling
library(tidymodels)         # Modeling framework (recipes, workflows, tune, yardstick, rsample)
library(themis)             # Class-imbalance sampling (e.g., SMOTE)
library(dials)              # Parameter tuning grids
library(ranger)             # Random forest engine
library(xgboost)            # Gradient boosting engine

# Goodness-of-fit & diagnostics
library(generalhoslem)      # Hosmer–Lemeshow tests
library(ResourceSelection)  # Additional GOF tests
library(rms)                # Regression modeling strategies
library(car)                # Companion to applied regression
